### 2024-03-29 parts/intro.tex
appendix + proofrefs

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\kl{$\Nat$-rational series} that was frequently rediscovered in the literature
\cite{SCHU62,KRRC13}. In \cite{CDTL23}, the authors introduced the class of
\kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$) that generalize
\kl{$\Nat$-polyregular functions} by allowing negative outputs. This remains a
subclass of \kl{rational series}, more precisely, of the class of
\kl{$\Rel$-rational series} ($\intro*\ZRat$). For this class of functions the
aperiodic variant $\intro*\ZSF$ has been shown to be decidable \cite{CDTL23}.
Note however that $\ZPoly$ does not correspond to a restriction of
\kl{polyregular functions}, because of the extra expressive power introduced by
the negative outputs \cite{CDTL23}. Furthermore, deciding the membership of
$\NRat$ inside $\ZRat$ is an open problem
(see e.g. \cite{KARH77}),
that remains open even in the
restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}.

\AP It was conjectured that, analogously to classes appearing in
\cref{computational-models:fig}, the following equation would hold: $\NSF =
\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
decidability of aperiodicity for single letter output polyregular functions
since aperiodicity is decidable for $\ZSF$. We refer the reader to
\cref{previously-known-inclusions:fig} to get an overview of the different
classes of functions appearing in the paper.

>>>>

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\kl{$\Nat$-rational series} that was frequently rediscovered in the literature
\cite{SCHU62,KRRC13}. In \cite{CDTL23}, the authors introduced the class of
\kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$) that generalize
\kl{$\Nat$-polyregular functions} by allowing negative outputs. This remains a
subclass of \kl{rational series}, more precisely, of the class of
\kl{$\Rel$-rational series} ($\intro*\ZRat$). For this class of functions the
aperiodic variant $\intro*\ZSF$ has been shown to be decidable \cite{CDTL23}.
Beware that concurrent (non-equivalent) notions of aperiodicity exists in the
research area of rational series \cite{REUT80,DRGA19,CDTL23}, and that we only
refer to the one applying to rational series of polynomial growths introduced
in \cite{CDTL23}.

\AP This approach through rational series has several limitations. First,
$\ZPoly$ does not correspond to a restriction of \kl{polyregular functions},
because of the extra expressive power introduced by the negative outputs
\cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
is an open problem (see e.g. \cite{KARH77}), that remains open even in the
restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It
was conjectured that, analogously to classes appearing in
\cref{computational-models:fig}, the following equation would hold: $\NSF =
\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
decidability of aperiodicity for single letter output polyregular functions
since aperiodicity is decidable for $\ZSF$. We refer the reader to
\cref{previously-known-inclusions:fig} to get an overview of the different
classes of functions appearing in the paper.

### 2024-03-29 parts/prelim.tex
appendix + proofrefs

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
\end{example}
\begin{proof}
    Let us define $M \defined (\set{1}, \times)$ which is 
    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
    defined by $\mu(w) \defined 1$, and
    $\pi \colon M^2 \to \Nat$
    that is the constant function equal to $1$.
    We check that for all $w \in \Sigma^*$:
    \begin{equation*}
        \pi^\dagger(w)
        =
        \sum_{uv = w} 1
        =
        \card{w} + 1
        = f(w)
        \quad . 
        \qedhere
    \end{equation*}
\end{proof}

@@

Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.

@@

\begin{example}
    \label{regular-language:ex}
    Let $\Sigma$ be a finite alphabet, and
    $L \subseteq \Sigma^*$. Then,
    $L$ is a regular language if and only if
    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
    \kl{$\Nat$-polyregular function}.
    Furthermore, $L$ is a star-free regular language
    if and only if $\ind{L}$ is a
    \kl{star-free $\Nat$-polyregular function}.
\end{example}

It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
of first order free variables used to construct the function as in
\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.

@@

\begin{example}
    \label{negative-not-nrat:ex}
    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomials},
    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
    not an \kl{$\Nat$-rational polynomial}.
\end{example}

>>>>

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
    \proofref{size-of-word-nsf:ex}
\end{example}

@@

@@

%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
%\begin{example}
%    \label{regular-language:ex}
%    Let $\Sigma$ be a finite alphabet, and
%    $L \subseteq \Sigma^*$. Then,
%    $L$ is a regular language if and only if
%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
%    \kl{$\Nat$-polyregular function}.
%    Furthermore, $L$ is a star-free regular language
%    if and only if $\ind{L}$ is a
%    \kl{star-free $\Nat$-polyregular function}.
%\end{example}
%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
%of first order free variables used to construct the function as in
%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.

@@

\begin{example}
    \label{negative-not-nrat:ex}
    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomials},
    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
    not an \kl{$\Nat$-rational polynomial}.
    \proofref{negative-not-nrat:ex}
\end{example}

### 2024-03-29 parts/starfree.tex
appendix + proofrefs

In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
of \emph{polynomial pumping arguments}
\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, we will provide a
semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

@@

>>>>

In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
of \emph{polynomial pumping arguments}
\cite{DOUE21,DOUE22,CDTL23}, we will provide a
semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

@@

\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}
relies on the construction of a canonical object called the \emph{residual
transducer}, the latter being essentially based on differences between
functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
proof method does not carry from $\Rel$-output functions to $\Nat$-output
functions, it was conjectured that the semantic property of being
\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$, a
conjecture that is restated hereafter.

\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
    \label{zsf-nsf:conjecture}
    Let $k \in \Nat$.
    A function $f \in \NPoly[k]$
    belongs to $\NSF[k]$ if and only if
    it is \kl{ultimately polynomial}.
    In particular,
    $\NSF[k] = \ZSF[k] \cap \NPoly$.
\end{conjecture}

We answer positively to \cref{zsf-nsf:conjecture} in the commutative case, by
leveraging the semantic characterizations respectively of $\ZSF$ inside
$\ZPoly$ (\kl{ultimately polynomial}) and $\NPoly$ inside $\ZPoly$
(\kl{$(k,\Nat)$-combinatorial}), which is possible thanks to the decomposition
obtained in \cref{decompose-polynomial:lem}. Decidability then follows
immediately from the effective conversions, and the previous decidability
result for $\ZSF$.

### 2024-03-29 parts/transducers.tex
appendix + proofrefs

Given the syntax of an \kl{$\mathcal{H}$-transducer}, let us now 
introduce its semantics, which is given by a simple recursion
scheme. Intuitively, the function computed by a transducer is

@@

\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
        \item $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
            (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
            \kl{counter-free}).
    \end{itemize}
\end{lemma}

@@

>>>>

@@

\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    Then, 
    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
        \kl{counter-free}).
\end{lemma}

@@

\AP
We can prove that for functions in $\NPoly[0]$, aperiodicity is equivalent to
the absence of counters in the \kl{residual transducer} of the function (see
\cref{aperiodic-iff-residual:lem}), however this result does not generalize to
higher growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number of
states computing the function is \kl{counter-free}, which suggests that a more
sophisticated object is needed to observe aperiodicity. In the spirit of the
characterization of $\NPoly$ based the $(\resleq{f}{k})$ quasi-ordering
(\cref{non-commutative-npoly:thm}), we introduce the notion of \intro{aperiodic
ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is \reintro[aperiodic
ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$, there exists $N_0
\in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$ is increasing.
This leads to the following \cref{sf-no-periods-on-sequences:conj}, of which
the easy implication is already known to hold
(\cref{sf-no-periods-on-sequences:lemma}), and requires the introduction of the
\emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of $(\resleq{f}{k})$,
defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u})
\in \NSF[k-1]$.

\begin{lemma}
    \label{aperiodic-iff-residual:lem}
    Let $f \in \NPoly[0]$. Then,
    $f \in \NSF$ if and only if 
    $f \in \ZSF$, if and only if
    $f$ is \kl{ultimately polynomial}, if and only if 
    the \kl{$0$-residual transducer} of $f$ is \kl{counter-free}.
\end{lemma}

\begin{example}
    \label{non-aperiodic-residual-transd:ex}
    Let us define
    $\BadExKo(\varepsilon) = 1$,
    $\BadExKo(a) = 0$,
    $\BadExKo(a^2) = 1$,
    and $\BadExKo(a^n) = n - 3$ for all $n \geq 3$.
    The \kl{$0$-residual transducer} of $\BadExKo$ has a \kl{counter} and two states.
    Furthermore,
    every \kl{$\NPoly[0]$-transducer} with two states contains a \kl{counter}.
\end{example}

\begin{lemma}
    \label{sf-no-periods-on-sequences:lemma}
    Let $k \in \Nat$, and $f \in \SF[k]$. Then,
    $(\Sigma^*, \resleqsf{f}{k})$ is an
    \kl[aperiodic ordering]{aperiodic} \kl{well-quasi-ordering}.
\end{lemma}

\begin{conjecture}
    \label{sf-no-periods-on-sequences:conj}
    For all $k \in \Nat$ and $f \colon \Sigma^* \to \Rel$,
    $f \in \NSF[k]$ if and only if $(\Sigma^*, \resleqsf{f}{k})$ is an \kl[aperiodic ordering]{aperiodic}
    \kl{well-quasi-ordering}.
\end{conjecture}
